Discrete Time Markov Model

### 2025-03-28 analysis/data-overview.qmd
Updating site

---
title: "Procrastination & Dementia: A Markov Model"
author: 
  - name: "Cormac Monaghan"
    url: "https://c-monaghan.github.io/"
    orcid: "https://orcid.org/0000-0001-9012-3060"
    affiliation:
      - "Hamilton Institute, Maynooth University, Ireland"
      - "Department of Psychology, Maynooth University, Ireland"
  - name: "Rafael de Andrade Moral"
    url: "https://rafamoral.github.io/"
    orcid: "https://orcid.org/0000-0002-0875-3563"
    affiliation:
      - "Department of Mathematics and Statistics, Maynooth University, Ireland"
  - name: "Joanna McHugh Power"
    orcid: "https://orcid.org/0000-0002-7387-3107"
    affiliation: 
      - "Department of Psychology, Maynooth University, Ireland"
toc: true
toc-depth: 4
toc-title: On this page
bibliography: references.bib
csl: apa.csl
format: 
  html:
    theme: 
      light: united
    page-layout: full
    html-math-method: katex
    code-tools: true
    self-contained: true
    code-fold: true
    code-summary: "Show the code"
    include-in-header: 
      - file: ./html/github-corner.html
include-after-body: ./html/footer.html
execute: 
  warning: false
editor: source
---

```{r setting-up}
rm(list = ls())

set.seed(43561)

# Packages
suppressMessages(library(dplyr))
library(tidyr)
library(ggplot2)
library(ggalluvial)

source(here::here("./02__Models/00__Functions.R"))

# Paths -----------------------------------------------------------------------
path_data <- "./01__Data/02__Processed/"
```

@@

![LWC Workflow](./images/LWC_Workflow.png){#fig-lwc}

# Dementia Dataset

Our updated dataset contains 13 columns of cognitive classifications spanning from 1996 to 2022. These columns reflect dementia status across each wave, using the same criteria and structure as the original Langa-Weir dataset, now extended to include the most recent 2022 wave.

```{r data-import}
data <- readxl::read_xlsx(here::here(path_data, "data.xlsx"))

dementia_data <- data |>
  count_transitions(years = seq(1996, 2022, by = 2))

glimpse(dementia_data)
```

## Missing data

Since the participant pool is drawn from the 2020 wave of the HRS data, any analysis that looks back at previous waves will inevitably encounter missing data. Not all participants were included in earlier waves, leading to gaps in the data. @fig-missing-data displays the frequency of missing cognitive test data across each HRS wave.

```{r missing-data}
#| fig-width: 8
#| label: fig-missing-data
#| fig-cap: Occurance of missing data across HRS waves (participants were gathered from the 2020 HRS wave [in red])

# Dynamic colours
colours <- c(rep("#2e2e2e", times = 12), "red", "#2e2e2e")

data |>
  select(any_of(paste0("cogtot27_imp", 1996:2022))) |>
  rename_with( ~ stringr::str_replace(., "cogtot27_imp", ""), starts_with("cogtot27")) |>
  visdat::vis_dat(palette = "cb_safe") +
  labs(title = "Missing data across HRS Waves (1996 - 2022)",
       subtitle = "Total Cognitive Scores") +
  theme(
    plot.title = element_text(hjust = 0.5, size = 12, face = "bold", colour = "#2e2e2e"),
    plot.subtitle = element_text(hjust = 0.5, size = 10, colour = "#2e2e2e"),
    axis.text.x = element_text(colour = colours)) +
  ggeasy::easy_remove_legend() +
  ggeasy::easy_x_axis_labels_size(size = 9)

```

The HRS had its most recent participant intake in 2016, which explains the notable decline in missing data occurrences from that point onward. As new participants were not added after 2016, the dataset becomes more complete in subsequent waves, with fewer missing values.

Given this shift, we will conduct our analysis **the reduced dataset (2016 - 2022)** (@fig-complete-data) along with removing the missing values from 2016 and 2018.

```{r complete-data}
#| fig-width: 8
#| label: fig-complete-data
#| fig-cap: HRS dataset after processing

data |>
  select(any_of(paste0("cogtot27_imp", 2016:2022))) |>
  rename_with( ~ stringr::str_replace(., "cogtot27_imp", ""), starts_with("cogtot27")) |>
  na.omit() |>
  visdat::vis_dat(palette = "cb_safe") +
  labs(title = "Complete dataset across HRS Waves (2016 - 2022)",
       subtitle = "Total Cognitive Scores") +
  theme(
    plot.title = element_text(hjust = 0.5, size = 12, face = "bold", colour = "#2e2e2e"),
    plot.subtitle = element_text(hjust = 0.5, size = 10, colour = "#2e2e2e")) +
  ggeasy::easy_remove_legend() +
  ggeasy::easy_x_axis_labels_size(size = 9)
```

## Exploratory Data Analysis

### Cognitive Test Scores

Initially, we will plot a distribution of the cognitive test scores (ranging from 0 - 27) across time for all HRS participants.

```{r cog-scores}
#| fig-width: 12
#| label: fig-cognitive-test-scores
#| fig-cap: Cognitive test scores (2016 - 2022)

plot_cognitive_scores(data = data, year = 2016)
```

### Classification proportion per year

@fig-dementia-proportions shows the proportion of dementia classifications from HRS 2016 - 2022. The variable **Missing** represents the procrastination HRS respondents who were not yet interviewed by the HRS. Since the analysis focuses on participants included in the 2020 HRS wave, any retroactive analysis of prior waves may result in missing data for certain individuals

```{r cog-class}
#| fig-width: 12
#| label: fig-dementia-proportions
#| fig-cap: Proportion of dementia classifications per year

# Plotting proportions --------------------------------------------------------
data |>
  count_transitions(years = seq(2016, 2022, by = 2), absorbing = TRUE) |>
  ggplot(aes(x = Wave, fill = Classification)) +
  geom_bar(position = "fill", alpha = 0.5, colour = "black") +
  scale_y_continuous(labels = scales::percent) +
  labs(title = "Proportion of Cognitive Status Classifications (2016 - 2022)", 
       x = "", y = "Percentage") +
  scale_fill_viridis_d(direction = -1) +
  theme_minimal() +
  theme(panel.grid = element_blank(),
        plot.title = element_text(size = 14, hjust = 0.5, face = "bold"),
        axis.text = element_text(size = 10),
        axis.title = element_text(size = 10),
        text = element_text(size = 10)) +
  ggeasy::easy_move_legend("bottom") +
  ggeasy::easy_remove_legend_title()
```

### Dementia transitions per year

@fig-dementia-transitions is an alluvial graph illustrating the transitions in cognitive classifications from one HRS wave to the next.

```{r cog-transition}
#| fig-width: 12
#| label: fig-dementia-transitions
#| fig-cap: Dementia transitions (2016 - 2022)

data |>
  count_transitions(years = seq(2016, 2022, by = 2)) |>
  group_by(ID, Wave) |>
  reframe(plyr::count(Classification)) |>
  rename(Classification = x, n = freq) |>
  plot_transitions(size = 2.5)
```

From this figure we can see that some individuals classified with dementia in one wave transition out of dementia in a subsequent wave. Let's fix this by adding an additional parameter to our `count_transitions()` function to turn dementia into an absorbing state (@fig-dementia-transitions-absorbing). We can do this by using the `cumany()` function from the `dplyr` [@Wickham2023] package.

```{r cog-transition-absorbing}
#| fig-width: 12
#| label: fig-dementia-transitions-absorbing
#| fig-cap: Dementia transitions with dementia as an absorbing state (2016 - 2022)

data |>
  count_transitions(years = seq(2016, 2022, by = 2), absorbing = TRUE) |>
  group_by(ID, Wave) |>
  reframe(plyr::count(Classification)) |>
  rename(Classification = x, n = freq) |>
  plot_transitions(size = 2.5)
```

# Markov Modelling

In this section, we apply Markov modelling to analyze the transitions between cognitive states (Normal Cognition, Mild Cognitive Impairment [MCI], and Dementia) using longitudinal data from the Health and Retirement Study (HRS) from 2016 to 2022. By modeling these transitions, we aim to understand the probabilities of moving between different cognitive states over time.

## Data preparation

The first step is to prepare the transition dataset. We start by transforming the data to create a record of transitions from one cognitive state to another between consecutive waves.

Once the transitions are identified, we calculate the probabilities of each state-to-state transition. These probabilities are computed by counting the occurrences of each transition and dividing by the total number of transitions.

```{r transition-probability}
tran_prob <- data |>
  extract_years(years = seq(2016, 2022, by = 2)) |>
  create_transitions(absorbing = TRUE) |>
  calculate_probabilties()

tran_prob |>
  mutate(prop = round(prop, digits = 3)) |>
  head(n = 9) |>
  gt::gt() |>
  gtExtras::gt_theme_538() |>
  gt::cols_align(align = "center") |>
  gt::tab_header(title = "Overall Transition Probabilties (2016 - 2022)") |>
  gt::tab_options(table.width = gt::pct(100)) |>
  gt::opt_align_table_header(align = "center")
```

## Creating transition matrix

The probability distribution of transitions from one state to another can be represented into a transition matrix $P = (p_{ij})_{i,j}$ where each element of position $(i, j)$ represents the transition probability $p_{ij}$.

$$
P = \begin{bmatrix} 
p_{11} & p_{12} & p_{13} \\
p_{21} & p_{22} & p_{23} \\
p_{31} & p_{32} & p_{33} \\
\end{bmatrix}
$$

```{r transition-matrix}
tran_matrix <- tran_prob |>
  transition_matrix() |>
  normalise() |>
  round(digits = 3)

tran_matrix
```

This code creates a $3 \times 3$ matrix (for the three states: Normal Cognition, MCI, and Dementia) and populates it with the transition probabilities.

## Visualising the matrix

To make the transition matrix more interpretable we visualize the probabilities using a heat map (@fig-transition-matrix)

```{r transition-matrix-plot}
#| fig-width: 8
#| label: fig-transition-matrix
#| fig-cap: Heat map of transition probabilities (2016 - 2022)

plot_matrix(data = t(tran_matrix), year = 2016)
```

We can also create and plot transition matrices for each wave separately allowing us to see how these transition probabilities change across multiple waves of data collection

```{r transition-matrix-longitudinal}
tran_prob_long <- data |>
  extract_years(years = seq(2016, 2022, by = 2)) |>
  create_transitions(absorbing = TRUE) |>
  group_by(Wave) |>
  calculate_probabilties() |>
  group_map(~ transition_matrix(.x, longitudinal = TRUE), .keep = TRUE) |>
  lapply(function(mat) {
    mat[c("Normal Cognition", "MCI", "Dementia"), c("Normal Cognition", "MCI", "Dementia")] |>
        normalise()
    }) |>
  purrr::map2_df(.y = c(seq(2016, 2020, by = 2)), .f = reshape_matrix) |>
  mutate(
    transition_prob = round(transition_prob, digits = 3),
    from = factor(from, levels = c("Normal Cognition", "MCI", "Dementia"))
  )

DT::datatable(
  tran_prob_long,
  filter = "top",
  colnames = c("Probability" = "transition_prob"),
  options = list(pageLength = 6, autoWidth = TRUE)
)
```

```{r transition-matrix-longitudinal-plot, fig.width=8, fig.show="animate", ffmpeg.format="gif", dev="png"}
for(wave in c(seq(2016, 2020, by = 2))){
  print(plot_transition_wave(data = tran_prob_long, wave = wave))
}
```

# Multi-State Markov Model

Now that we have an idea of the disease progression across time we are interested in the effect of different covariates on these transitions. For this, we can fit a multi-state Markov model. Multi-state models (MSMs) provide a comprehensive view of the disease process and make risk-factors analysis and long-term predictions more easily. The conditional distribution of the status of an individual participant at an arbitrary examination given their status at previous examinations was assumed to have the **Markov property**. That is the distribution of the forthcoming state $X_{n+1}$ depends only on the current state $X_n$ and doesn't depend on the previous ones $X_{n-1}, X_{n-2}, \dots, X_1$.

$$Pr(X_{n+1} = x_{n+1} \; \vert \; X_1 = x_1, X_2 = x_2, \dots, X_n = x_n) = Pr(X_{n+1} = x_{n+1} \; \vert \; X_n = x_n)$$

## Data preperation

For our model we are going to look at several different covariates:

- **Gender**: $0 = \text{Male;} \; 1 = \text{Female}$ 
- **Age**: $\text{Range} \; = 50 - 91$ 
- **Education**: $0 = \text{No Degree;} \; 1 = \text{High School Degree;} \; 2 = \text{Further Education}$ 
- **Number of cardiovascular risk factors**: $\text{Range} \; = 0 - 4$
- **Procrastination**: $\text{Range} \; = 4 - 60$

```{r markov-prep}
cols <- c(
  "ID", "Gender", "Age", "Education_tri", 
  paste0("Cardio_risk_", seq(16, 22, by = 2)), 
  paste0("Total_dep_20", seq(16, 22, by = 2)), "Total_p")

markov_data <- data |>
  extract_years(seq(2016, 2022, by = 2)) |>
  na.omit() |>
  inner_join(data[, cols], by = "ID") |>
  pivot_data() |>
  backtrack_age(base_year = 2016, current_year = 2022) |>
  group_by(ID) |>
  mutate(
    Age = min(Age), # Baseline age
    Cardio_risk = Cardio_risk[Wave == 1],
    Depression = Depression[Wave == 1]
    ) |>
  ungroup() |>
  filter(Age >= 50) # Only looking at older adults

DT::datatable(
  markov_data,
  filter = "top",
  options = list(pageLength = 6, autoWidth = TRUE)
)
```

Let's look at a state table to visualize the amount of transitions over time

```{r state-table}
markov_data |>
  select(ID, Wave, Status) |>
  mutate(Status = case_when(
    Status == 1 ~ "Normal Cognition", 
    Status == 2 ~ "MCI", 
    Status == 3 ~ "Dementia")) |>
  msm::statetable.msm(state = Status, subject = ID)
```

## Modelling

### Defining a q-matrix

The Q matrix (also known as the transition intensity matrix or generator matrix) defines the rates at which transitions occur between states in the model.

It is a square matrix where:

- Each row corresponds to a **from** state in the model.
- Each column corresponds to a **to** state in the model.
- The off-diagonal elements $q_{ij}$ represent the transition rate from state $i$ to state $j$.

For a model with $n$ states, the Q matrix has the following structure

$$
Q = \begin{bmatrix} 
q_{11} & q_{12} & \cdots & q_{1n} \\
q_{21} & q_{22} & \cdots & q_{2n} \\
\vdots & \vdots & \ddots &\vdots  \\
q_{n1} & q_{n2} & \cdots & q_{nn}
\end{bmatrix}
$$

```{r q-matrix}
# Creating transition matrix
# Define possible transitions
states <- c("Normal Cognition", "MCI", "Dementia")

# Theoretical transition probabilities
q_matrix <- rbind(
  c(0.75, 0.20, 0.05),  # From Normal, can transition to MCI or Dementia
  c(0.25, 0.50, 0.25),  # From MCI, can transition to Normal or Dementia
  c(0.00, 0.00, 1.00)         # From Dementia, no transitions (absorbing state)
)

rownames(q_matrix) <- states
colnames(q_matrix) <- states

q_matrix
```

## Fitting

We will fit three different multi-state models

1. A model with no covariates
2. A multivariate model with covariates acting on all transitions
3. A multivariate model with covariates (and interactions) acting on all transitions
4. A multivariate model with covariates acting on only NC <-> MCI & MCI - Dementia
5. A multivariate model with covariates (and interactions) acting on only NC <-> MCI & MCI - Dementia

```{r model-fitting}
# Fitting model with no covariates
model_1 <- msm::msm(
  formula = Status ~ Wave,
  subject = ID,
  data = markov_data,
  qmatrix = q_matrix,
  obstype = 1,
  method = "BFGS", 
  gen.inits = TRUE,
  control = list(
    fnscale = 5000, 
    maxit = 2000 # To reach convergence
  ))

# Multivariate model: covariates acting on all transitions
model_2 <- msm::msm(
  formula = Status ~ Wave,
  subject = ID,
  data = markov_data,
  qmatrix = q_matrix,
  covariates = ~ Gender + Age + Education_tri + Cardio_risk + Depression + Total_p,
  obstype = 1,
  gen.inits = TRUE,
  method = "BFGS",
  control = list(
    fnscale = 5000, 
    maxit = 2000 # To reach convergence
  ))

# Multivariate model: covariates acting on all transitions (with interaction)
model_3 <- msm::msm(
  formula = Status ~ Wave,
  subject = ID,
  data = markov_data,
  qmatrix = q_matrix,
  covariates = ~ Gender + Age + Education_tri + Cardio_risk + (Depression * Total_p),
  obstype = 1,
  gen.inits = TRUE,
  method = "BFGS",
  control = list(
    fnscale = 5000, 
    maxit = 2000 # To reach convergence
  ))

# Multivariate model: covariates acting on NC <-> MCI and MCI -> D
model_4 <- msm::msm(
  formula = Status ~ Wave,
  subject = ID,
  data = markov_data,
  qmatrix = q_matrix,
  covariates = list(
    "1-2" = ~ Gender + Age + Education_tri + Cardio_risk + Depression + Total_p,
    "2-1" = ~ Gender + Age + Education_tri + Cardio_risk + Depression + Total_p,
    "2-3" = ~ Gender + Age + Education_tri + Cardio_risk + Depression + Total_p
  ),
  obstype = 1,
  gen.inits = TRUE,
  method = "BFGS",
  control = list(
    fnscale = 5000, 
    maxit = 2000 # To reach convergence
  ))

# Multivariate model: covariates acting on NC <-> MCI and MCI -> D (with interaction)
model_5 <- msm::msm(
  formula = Status ~ Wave,
  subject = ID,
  data = markov_data,
  qmatrix = q_matrix,
  covariates = list(
    "1-2" = ~ Gender + Age + Education_tri + Cardio_risk + (Depression * Total_p),
    "2-1" = ~ Gender + Age + Education_tri + Cardio_risk + (Depression * Total_p),
    "2-3" = ~ Gender + Age + Education_tri + Cardio_risk + (Depression * Total_p)
  ),
  obstype = 1,
  gen.inits = TRUE,
  method = "BFGS",
  control = list(
    fnscale = 5000, 
    maxit = 20002 # To reach convergence
  ))

```

### Model Checking

After fitting each of our models we can extract some goodness of fit criteria

```{r model-goodness}
log_l <- c(logLik(model_1), 
           logLik(model_2), 
           logLik(model_3), 
           logLik(model_4), 
           logLik(model_5))

aic <- AIC(model_1, model_2, model_3, model_4, model_5)

# Very small difference between model 2 and 3 (makes sense)
cbind(log_l, aic) |>
  tibble::rownames_to_column("Model") |>
  mutate(Model = stringr::str_remove(Model, "model_"))

# Likelihood ratio test
lmtest::lrtest(model_1, model_2, model_3, model_4, model_5)
```

We can also look at a measure of how well the model fits the data using a test proposed by [@gentleman1994]. The function plots the observed numbers of individuals occupying a state at a series of times against forecasts from the fitted model, for each state (@fig-gentleman).

```{r gentleman_test}
#| fig-width: 12
#| fig-height: 12
#| label: fig-gentleman
#| fig-cap: Fitted values versus observed values

# Plotting fitted vs. observed values
g1 <- gentleman_test(data = msm::prevalence.msm(model_1), 
                     model = "Model 1")
g2 <- gentleman_test(data = msm::prevalence.msm(model_2), 
                     model = "Model 2")
g3 <- gentleman_test(data = msm::prevalence.msm(model_3), 
                     model = "Model 3")
g4 <- gentleman_test(data = msm::prevalence.msm(model_4), 
                     model = "Model 4")
g5 <- gentleman_test(data = msm::prevalence.msm(model_5), 
                     model = "Model 5")

ggpubr::ggarrange(g1, g2, g3, g4, g5, 
                  common.legend = TRUE, 
                  nrow = 5,
                  heights = c(2, 2, 2, 2, 2), 
                  legend = "bottom")
```

### Hazard ratios

We now present hazard ratios for each of our fitted models. These ratios represent how the instantaneous risk of making a particular transition is modified by the covariate.

::: {.panel-tabset}

## Model 2

@fig-hazard-ratio-1 presents hazard ratios from model 2 (our multivariate model with covariates acting on all transitions) for each covariate.

```{r}
#| fig-width: 12
#| fig-height: 12
#| label: fig-hazard-ratio-1
#| fig-cap: Ouputted hazard ratios from Markov model. Green bars represent a decreased risk of transition while red bars indicate an increased risk of transition.

# Plotting hazard ratios
model_2 |>
  msm::hazard.msm() |>
  process_hr() |>
  calculate_p_value() |>
  plot_hr()
```

## Model 3

@fig-hazard-ratio-2 presents hazard ratios from model 3 (our multivariate model with covariates **(and interactions)** acting on all transitions) for each covariate.

```{r}
#| fig-width: 12
#| fig-height: 12
#| label: fig-hazard-ratio-2
#| fig-cap: Ouputted hazard ratios from Markov model. Green bars represent a decreased risk of transition while red bars indicate an increased risk of transition.

# Plotting hazard ratios
model_3 |>
  msm::hazard.msm() |>
  process_hr() |>
  calculate_p_value() |>
  plot_hr()
```

## Model 4

@fig-hazard-ratio-3 presents hazard ratios for model 4 (our multivariate model with covariates acting on only NC <-> MCI & MCI -> Dementia) for each covariate

```{r}
#| fig-width: 12
#| fig-height: 12
#| label: fig-hazard-ratio-3
#| fig-cap: Ouputted hazard ratios from Markov model. Green bars represent a decreased risk of transition while red bars indicate an increased risk of transition.

# Plotting hazard ratios
model_4 |>
  msm::hazard.msm() |>
  process_hr() |>
  calculate_p_value() |>
  plot_hr()
```

## Model 5

@fig-hazard-ratio-4 presents hazard ratios for model 5 (our multivariate model with covariates **(and interactions)** acting on only NC <-> MCI & MCI -> Dementia) for each covariate

```{r}
#| fig-width: 12
#| fig-height: 12
#| label: fig-hazard-ratio-4
#| fig-cap: Ouputted hazard ratios from Markov model. Green bars represent a decreased risk of transition while red bars indicate an increased risk of transition.

# Plotting hazard ratios
model_5 |>
  msm::hazard.msm() |>
  process_hr() |>
  calculate_p_value() |>
  plot_hr()
```

:::

>>>>

---
title: "Data Overview"
format: 
  html:
    code-tools: false
---

@@

![LWC Workflow](../files/workflow.svg){#fig-lwc}

### 2025-03-28 analysis/discrete-markov.qmd
Updating site

---
title: "Discrete Time Markov Model Analysis of Cognitive Decline"
subtitle: "Examining the Association Between Procrastination and Cognitive Transitions"
author: 
  - name: "Cormac Monaghan"
    url: "https://c-monaghan.github.io/"
    orcid: "https://orcid.org/0000-0001-9012-3060"
    affiliation:
      - "Hamilton Institute, Maynooth University, Ireland"
      - "Department of Psychology, Maynooth University, Ireland"
abstract: |
  This analysis examines transitions between cognitive states (Normal Cognition, MCI, Dementia) using discrete-time Markov models, with particular focus on procrastination as a risk factor.
toc: true
toc-depth: 3
toc-title: On this page
number-sections: true
format: 
  html:
    theme: 
      light: cosmo
    page-layout: full
    shift-heading-level-by: 1
    html-math-method: katex
    code-tools: true
    code-copy: true
    code-block-bg: "#333A3F"
    highlight-style: monokai
    fig-align: center
    self-contained: true
---

>>>>

---
title: "Discrete Time Markov Model"
subtitle: "Examining the Association Between Procrastination and Cognitive Transitions"
---
